\documentclass[11pt]{article}

% ---------------------------------------------------------------------------
% Packages
% ---------------------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds,calc}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage{caption}
\usepackage{subcaption}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{authblk}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

% ---------------------------------------------------------------------------
% Listing style
% ---------------------------------------------------------------------------
\definecolor{codebg}{rgb}{0.97,0.97,0.97}
\lstset{
  basicstyle=\ttfamily\footnotesize,
  backgroundcolor=\color{codebg},
  breaklines=true,
  frame=single,
  framesep=4pt,
  rulecolor=\color{gray!40},
  showstringspaces=false,
  columns=fullflexible,
}

\newcommand{\sysname}{MemoPilot~IQ}
\newcommand{\osname}{MemoryOS}
\newcommand{\cmark}{\textcolor{green!55!black}{\checkmark}}
\newcommand{\xmark}{\textcolor{red!65!black}{\ensuremath{\times}}}
\newcommand{\pmark}{\textcolor{orange!75!black}{$\circ$}}

\title{\textbf{\sysname{}: A Transparent, Budget-Aware Memory Layer\\
for Persistent Large Language Model Agents}}

\author[1]{Mahar Muavia\thanks{Corresponding author: \texttt{contact@inventacore.org}}}
\author[1,*]{Muhammad Imran Shah}
\author[1,*]{Tayyaba Shehzad}
\affil[1]{InventaCore AI \quad \url{https://inventacore.org}}
\affil[*]{These authors contributed equally.}

\date{\today}

% ---------------------------------------------------------------------------
\begin{document}
\maketitle

\begin{center}
\fcolorbox{red!65!black}{red!5}{\parbox{0.93\linewidth}{\textbf{Draft status.}
This is an archived preliminary research draft. Its empirical tables and
numeric claims predate the final evaluator and strict-budget fixes, so they
are not submission evidence and must not be quoted. Regenerate every result
from the current harnesses before publication.}}
\end{center}

\begin{abstract}
Large language model (LLM) assistants are stateless across sessions: they forget
user preferences, project decisions, and constraints, and continue to give
\emph{outdated} advice after a decision has changed. We present \sysname{}, a
persistent-memory agent organized around \osname{}, a self-contained memory
intelligence layer that sits between the conversation and the LLM. \osname{}
(i)~extracts \emph{typed, structured} memories from dialogue rather than storing
raw transcripts; (ii)~ranks candidates with a single, fully interpretable
linear scoring function that fuses semantic similarity with intrinsic signals
(importance, recency, confidence, usage), contextual signals (project match,
criticality), and explicit penalties (outdated, privacy, supersession);
(iii)~assembles the prompt under a strict token budget that prioritizes
critical and pinned memories when they fit; (iv)~performs non-destructive
forgetting through expiry, archival, and supersession state transitions; and
(v)~emits a complete \emph{Memory Trace} that records, for every answer, which
memories were injected or skipped, their component scores, the reason, and the
token cost. We formalize the scoring and budgeting procedures and describe a
reproducible benchmark harness that contrasts the memory-augmented agent against
a no-memory baseline across 24 diagnostic scenarios spanning six capability
categories: preference recall, cross-session recall, contradiction/supersession,
expired-deadline avoidance, critical-instruction retention, and multi-fact
composition. With a live \textsc{Qwen} backend (\texttt{qwen3.7-max}) the memory
agent reaches a task accuracy of $0.75$ versus $0.50$ for the no-memory baseline
within a $2{,}500$-token context budget, and the same gain holds when only the
answer model is swapped to GPT-4o ($0.79$ versus $0.67$), evidence that the
benefit comes from the memory layer rather than any single model. The layer
recalls target facts at \textsc{recall@5} $=0.77$ and reduces injected context
by roughly $98\%$ relative to replaying full history. A deterministic ablation
isolates the load-bearing mechanism: removing lifecycle exclusion raises the
outdated-memory leak rate ninefold with no recall benefit. On the standard
LoCoMo long-conversation benchmark, an evidence-conditioned analysis shows
answer F1 is $8.4\times$ higher when the memory layer places the annotated
evidence into context than when it does not, localizing quality to retrieval
depth rather than the answer model. We argue
that \emph{interpretability and lifecycle management}, not raw capacity, are the
decisive properties for trustworthy long-term agent memory, and we release the
full system, benchmark, and trace format to support reproducible study.
\end{abstract}

% ===========================================================================
\section{Introduction}
\label{sec:intro}

Transformer-based LLMs condition only on the tokens inside their context
window~\citep{vaswani2017attention,brown2020gpt3}. In a multi-session assistant
this produces a structural failure mode: once a conversation ends, everything
the user established is gone, including their stack preferences, the decisions
made for a project, hard constraints, deadlines, and past mistakes. Users are
forced to re-explain context at the start of every session, and, worse, the
assistant keeps repeating advice that a \emph{later} decision has already
overturned. A user who first chose React and later switched to Next.js will
keep receiving React advice, because the model has no notion that one statement
\emph{supersedes} the other.

Two families of remedies dominate practice. The first simply enlarges the
context window and replays long histories; but attention degrades over long
inputs~\citep{liu2024lostmiddle}, cost scales with length, and stale facts are
replayed alongside fresh ones with no mechanism to distinguish them. The second
is retrieval-augmented generation (RAG)~\citep{lewis2020rag}, which fetches
relevant passages from a vector store~\citep{karpukhin2020dpr,johnson2019faiss}.
RAG addresses \emph{capacity} but not \emph{lifecycle}: a vector store has no
opinion on whether a retrieved memory is still valid, whether it must always be
honored, or whether it should have been forgotten. It also typically offers no
account of \emph{why} a particular memory entered the prompt, an obstacle to
trust and debugging.

We take the position that the hard problems of agent memory are not capacity
problems but \emph{governance} problems: deciding what is worth remembering,
what must never be dropped, what has become invalid, and being able to
\emph{explain} every such decision. \sysname{} operationalizes this view through
\osname{}, a modular memory layer with five cooperating components (structured
extraction, interpretable scoring, hybrid retrieval, budget-aware assembly, and
non-destructive forgetting) unified by a transparent trace. The system runs in
two interchangeable modes: a fully offline local mode (SQLite with in-process
vectors and a deterministic embedding/LLM fallback) for reproducible evaluation,
and a cloud mode backed by managed object and table storage with Qwen
models~\citep{bai2023qwen,yang2024qwen2} for the language and embedding calls.

\paragraph{Contributions.}
\begin{enumerate}[leftmargin=1.4em,itemsep=2pt]
  \item \textbf{A typed memory schema and lifecycle.} A single record type with
  12 semantic categories, 6 lifecycle states, and three privacy levels that
  makes governance decisions first-class data rather than prompt heuristics
  (\S\ref{sec:schema}).
  \item \textbf{An interpretable scoring function.} A transparent linear model
  that fuses semantic, intrinsic, contextual, and penalty signals, with
  exponential recency decay and diminishing-returns usage scoring, so that every
  ranking decision decomposes into named, inspectable terms (\S\ref{sec:scoring}).
  \item \textbf{Budget-aware assembly with guaranteed criticality.} A context
  builder that fills a strict token budget by score while \emph{always}
  injecting critical/pinned memories, even when over budget, separating
  ``always honor'' instructions from ``recall if relevant'' ones
  (\S\ref{sec:budget}).
  \item \textbf{Non-destructive forgetting.} An engine that expires, archives,
  and supersedes memories through auditable state transitions, never silently
  hard-deleting user data (\S\ref{sec:forgetting}).
  \item \textbf{End-to-end transparency and evaluation.} A Memory Trace that
  exposes the full decision surface of every answer, together with a
  reproducible benchmark against a no-memory baseline across two frontier
  backbones, a deterministic ablation isolating the load-bearing mechanism, and
  a checkpointed LoCoMo adapter with a model-independent evidence-recall metric
  (\S\ref{sec:trace}--\S\ref{sec:locomo}).
\end{enumerate}

% ===========================================================================
\section{Related Work}
\label{sec:related}

\paragraph{Memory for LLM agents.}
Interest in equipping LLM agents with long-term memory has grown rapidly; recent
surveys catalogue the design space~\citep{wang2024memoryos}. \emph{Generative
Agents}~\citep{park2023generative} introduced a memory stream scored by a
weighted sum of recency, importance, and relevance, with periodic reflection
that synthesizes higher-level observations. \osname{}'s scoring function is in
the same spirit but extends the signal set with confidence, usage, project
match, and, critically, a set of explicit \emph{penalties} for outdated,
private, and superseded content, turning lifecycle state into a ranking signal.
MemGPT~\citep{packer2023memgpt} frames the LLM as an operating system that pages
information between a fixed context window and external storage; \sysname{}
shares the budget-management goal but emphasizes an interpretable, typed scoring
policy and a guaranteed-inclusion rule for critical content rather than learned
paging. MemoryBank~\citep{zhong2024memorybank} applies an
Ebbinghaus-style~\citep{ebbinghaus1885memory} forgetting curve to memory
strength; we instead make forgetting a discrete, auditable \emph{state machine}
(expire/archive/supersede) so that every transition is explainable and
reversible. A-Mem~\citep{xu2025amem} builds dynamically linked memory networks;
our Memory Graph provides a comparable relational view (supersession and
related-to edges) but is derived from explicit lifecycle metadata.

\paragraph{Production memory frameworks.}
A second, fast-moving line of work targets deployable memory layers.
Mem0~\citep{chhikara2025mem0} extracts facts and maintains them with
LLM-issued add/update/delete operations over a vector (and optional graph)
store, and reports strong long-conversation question-answering with low latency.
Zep~\citep{rasmussen2025zep} builds a \emph{temporal} knowledge graph
(Graphiti) with bi-temporal validity, invalidating facts as they are
contradicted over time. Letta~\citep{letta2024}, the productionized successor of
MemGPT, exposes self-editing memory blocks, and LangMem~\citep{langmem2025}
offers semantic/episodic/procedural memory tools for agent frameworks. Relative
to these, \sysname{} contributes orthogonally: rather than maximizing
long-context QA or temporal-graph expressiveness, it makes the \emph{selection
policy itself} interpretable and the lifecycle auditable. Mem0's update/delete
and Zep's temporal invalidation achieve supersession; \sysname{} differs in
exposing supersession as an inspectable state transition with a per-answer trace
and a guaranteed-inclusion rule for critical content. We report a first cut on
the shared LoCoMo benchmark~\citep{maharana2024locomo} in \S\ref{sec:locomo}; a
full, identically-graded head-to-head against Mem0 and Zep remains the most
important next experiment (\S\ref{sec:limitations}).

\paragraph{Retrieval.}
Dense retrieval~\citep{karpukhin2020dpr} and approximate nearest-neighbor
indexes~\citep{johnson2019faiss} underpin most RAG systems~\citep{lewis2020rag}.
Sparse lexical scoring such as BM25~\citep{robertson2009bm25} remains a strong
complement, particularly for rare tokens and exact identifiers. \osname{} uses a
lightweight hybrid of dense cosine similarity and sparse keyword/tag overlap,
which is robust precisely on the short, entity-heavy memories typical of
developer assistants (framework names, deployment targets, file paths).

\paragraph{Reasoning and self-improvement.}
Our reflection pass, which merges duplicates, promotes frequently used
memories, and derives higher-level insights, is related to
Reflexion~\citep{shinn2023reflexion} and to chain-of-thought / ReAct style
prompting~\citep{wei2022cot,yao2023react}, but operates over the persistent
memory store rather than over a single episode's trajectory.

\paragraph{Positioning.} Relative to prior systems, \sysname{}'s distinguishing
commitment is \emph{transparency by construction}: scoring is a named linear
combination, inclusion decisions carry human-readable reasons, forgetting is a
visible state machine, and a complete per-answer trace is a first-class API
object. This makes the memory layer auditable and debuggable, which we argue is
a precondition for deployment in settings where wrong but confident recall is
costly. Table~\ref{tab:comparison} contrasts these governance properties across
representative memory systems.

\begin{table}[t]
\centering
\small
\caption{Governance properties across representative LLM memory systems, by
primary design emphasis. \cmark~present, \pmark~partial, \xmark~not a focus.
\emph{Typed}: typed/structured records (vs.\ a raw stream); \emph{Interp.}:
inspectable scoring; \emph{Crit.}: guaranteed inclusion of critical items;
\emph{Audit.\ forget}: forgetting as an auditable state machine; \emph{Trace}:
per-answer inclusion/skip explanation.}
\label{tab:comparison}
\begin{tabular}{@{}lccccc@{}}
\toprule
\textbf{System} & \textbf{Typed} & \textbf{Interp.} & \textbf{Crit.}
  & \textbf{Audit.} & \textbf{Trace} \\
\midrule
Generative Agents~\citep{park2023generative} & \xmark & \pmark & \xmark & \xmark & \xmark \\
MemGPT / Letta~\citep{packer2023memgpt,letta2024} & \pmark & \xmark & \xmark & \xmark & \xmark \\
MemoryBank~\citep{zhong2024memorybank}       & \xmark & \pmark & \xmark & \pmark & \xmark \\
A-Mem~\citep{xu2025amem}                     & \pmark & \xmark & \xmark & \xmark & \xmark \\
Mem0~\citep{chhikara2025mem0}                & \cmark & \xmark & \xmark & \pmark & \xmark \\
Zep~\citep{rasmussen2025zep}                 & \cmark & \pmark & \xmark & \cmark & \pmark \\
\textbf{\sysname{}} (ours)                   & \cmark & \cmark & \cmark & \cmark & \cmark \\
\bottomrule
\end{tabular}
\end{table}

% ===========================================================================
\section{System Overview}
\label{sec:overview}

\sysname{} is a client/server application. A React/Vite frontend issues chat
and management requests to a FastAPI backend; all memory logic is encapsulated
in the \osname{} module so it can be embedded in other agents. Figure~\ref{fig:arch}
sketches the request lifecycle.

\input{figure_architecture}

\paragraph{Dual mode for reproducibility.} A central design choice is that the
entire system, including extraction, embeddings, retrieval, the benchmark, and
the test suite, runs with \emph{no external dependencies} via a deterministic
offline fallback for both embeddings and chat. This makes every
mechanism-level experiment in this paper (the ablation and all retrieval
diagnostics) exactly reproducible on a laptop with no credentials, and removes
nondeterminism from the evaluation of the memory \emph{mechanism} itself; the
live-model results are reproducible given an API key. In cloud mode the same
interfaces are backed by Qwen chat/embedding endpoints and managed
table/object storage; the active mode is reported on a health endpoint.

% ===========================================================================
\section{Memory Schema and Lifecycle}
\label{sec:schema}

The atomic unit is a \texttt{MemoryRecord}. Beyond content and an embedding
vector, each record carries the governance metadata that the rest of the system
reasons over (Table~\ref{tab:fields}).

\begin{table}[t]
\centering
\small
\caption{Selected fields of a \texttt{MemoryRecord}. Governance metadata
(type, status, importance, confidence, privacy, supersession links) is
first-class and drives scoring, retrieval, and forgetting.}
\label{tab:fields}
\begin{tabular}{@{}lp{0.62\linewidth}@{}}
\toprule
\textbf{Field} & \textbf{Role} \\
\midrule
\texttt{type} & One of 12 semantic categories (preference, project, decision,
mistake, constraint, deadline, learning\_goal, task, critical, temporary,
outdated, deleted\_by\_user). \\
\texttt{status} & Lifecycle state: active, pinned, archived, expired,
superseded, deleted. \\
\texttt{importance} & Intrinsic salience in $[0,1]$. \\
\texttt{confidence} & Extraction confidence in $[0,1]$. \\
\texttt{usage\_count} & Times the memory has been injected into context. \\
\texttt{recency} & Derived from \texttt{updated\_at} via exponential decay. \\
\texttt{privacy\_level} & public / private / sensitive. \\
\texttt{is\_critical} & ``Always honor'' flag (e.g. security constraints). \\
\texttt{supersedes / superseded\_by} & Directed links encoding decision
revisions. \\
\texttt{expires\_at} & Optional deadline for temporary memories. \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Structured extraction.} Rather than storing raw turns, a
``Memory Editor'' prompt asks the LLM to return JSON describing
\texttt{new\_memories}, \texttt{updates}, and \texttt{forget} actions, each typed
and scored for importance and confidence. Secrets (API keys, tokens) are
redacted before storage. This converts free-form dialogue into a normalized,
queryable store and is what allows downstream scoring and forgetting to operate
on semantics rather than text spans.

\paragraph{Lifecycle as a state machine.} A memory is born \texttt{active} (or
\texttt{pinned} if the user pins it). It may be \texttt{superseded} by a later,
contradicting decision; \texttt{expired} when its deadline passes;
\texttt{archived} when stale and unused; or \texttt{deleted} only on explicit
user request (Figure~\ref{fig:lifecycle}). Crucially, retrieval excludes
\{superseded, expired, deleted, archived\} from normal context while preserving
them on a timeline, so the agent never acts on invalid information yet the
history remains auditable.

\input{figure_lifecycle}

% ===========================================================================
\section{Interpretable Memory Scoring}
\label{sec:scoring}

Given a query $q$ with embedding $\mathbf{e}_q$ and a candidate memory $m$,
\osname{} computes a single scalar score as a weighted sum of named components.
Let $\mathrm{sim}(m,q)\in[0,1]$ be the hybrid similarity (\S\ref{sec:retrieval}),
$I_m,C_m\in[0,1]$ the importance and confidence, and define the derived signals

\begin{align}
\text{recency}(m) &= \exp\!\left(-\frac{\Delta t_m}{\tau}\right),
  \qquad \tau = 14~\text{days}, \label{eq:recency}\\
\text{usage}(m) &= 1 - \frac{1}{1 + \ln(1+u_m)}, \label{eq:usage}
\end{align}

where $\Delta t_m$ is the age in days since \texttt{updated\_at} and $u_m$ is the
usage count. Equation~\eqref{eq:recency} is an exponential forgetting curve with
a two-week half-life regime; Equation~\eqref{eq:usage} gives diminishing returns,
so a memory used many times is only modestly preferred over one used a few times.
Binary contextual indicators capture project match
($\text{proj}(m)=\mathbb{1}[\text{project}_m = \text{project}_q]$) and criticality
($\text{crit}(m)=\mathbb{1}[m~\text{is critical/pinned}]$). Penalty indicators flag
outdated content ($P^{\text{out}}_m$), privacy
($P^{\text{priv}}_m\in\{0,0.4,1\}$ for public/private/sensitive), and supersession
($P^{\text{sup}}_m=\mathbb{1}[\text{status}\in\{\text{superseded,expired,deleted}\}]$).
The final score is

\begin{equation}
\label{eq:score}
\begin{aligned}
S(m,q) =\ & 0.40\,\mathrm{sim}(m,q) + 0.20\,I_m + 0.15\,\text{recency}(m) \\
        & + 0.10\,C_m + 0.10\,\text{usage}(m) + 0.15\,\text{proj}(m) + 0.20\,\text{crit}(m) \\
        & - 0.30\,P^{\text{out}}_m - 0.25\,P^{\text{priv}}_m - 0.50\,P^{\text{sup}}_m.
\end{aligned}
\end{equation}

Three properties follow directly from the form of Equation~\eqref{eq:score} and
are central to the system's behavior:

\begin{description}[leftmargin=1.2em,itemsep=2pt]
  \item[Interpretability.] Every score decomposes into ten named terms. The
  trace (\S\ref{sec:trace}) reports each term, so a ranking can be explained and
  audited rather than merely trusted.
  \item[Semantics-led but not semantics-only.] Similarity is the largest single
  weight ($0.40$), so relevance dominates, but governance signals can override:
  a superseded memory pays a $-0.50$ penalty, larger than the full similarity
  weight, ensuring stale decisions sink below valid ones even when textually
  similar to the query.
  \item[Criticality as a soft-then-hard rule.] The $+0.20$ critical bonus raises
  critical memories in the ranking, and the budget stage (\S\ref{sec:budget})
  converts this into a \emph{hard} guarantee of inclusion.
\end{description}

The specific weights are hand-set priors chosen to encode these qualitative
commitments; we treat them as a transparent default policy and discuss
data-driven weight learning as future work (\S\ref{sec:limitations}).

% ===========================================================================
\section{Hybrid Retrieval}
\label{sec:retrieval}

For each candidate, \osname{} blends a dense and a sparse signal. The dense
signal is cosine similarity between the query and memory embeddings, clamped to
$[0,1]$. The sparse signal is the fraction of query tokens that appear in the
memory's content or tags,
\[
\text{kw}(m,q) = \frac{|\,T(q)\cap T(m)\,|}{|\,T(q)\,|},
\]
where $T(\cdot)$ denotes the lowercased alphanumeric token set. The combined
similarity is
\begin{equation}
\mathrm{sim}(m,q) = \max\!\Big(\text{dense}(m,q),\ \tfrac{1}{2}\text{dense}(m,q) + \tfrac{1}{2}\text{kw}(m,q)\Big),
\end{equation}
which never underperforms pure dense retrieval and boosts memories with exact
lexical overlap, valuable for the identifier-heavy content of developer
assistants. Retrieval first applies structured filters (user, project, and the
exclusion of inactive states), scores all remaining candidates with
Equation~\eqref{eq:score}, and returns them sorted so that critical/pinned
records lead, then by score. Because memory stores per user/project are small
relative to web-scale corpora, exhaustive scoring is fast (sub-millisecond to
low-millisecond in local mode); an ANN index~\citep{johnson2019faiss} can be
substituted transparently behind the same interface at scale.

% ===========================================================================
\section{Budget-Aware Context Assembly}
\label{sec:budget}

Retrieved candidates are not all injected. A context builder fills a strict
token budget $B$ (default $B=2{,}500$) under the following policy, with token
cost estimated at $\approx 4$ characters/token:

\begin{enumerate}[leftmargin=1.4em,itemsep=1pt]
  \item Order candidates with critical/pinned first, then by score $S(m,q)$.
  \item Inject critical/pinned memories \emph{unconditionally}, even if doing so
  exceeds $B$ (the budget is overridden, and the trace records this).
  \item Inject the top-$k$ ($k=8$) remaining memories in score order while they
  fit within $B$.
  \item Skip the rest, recording for each a reason, either ``token budget
  exhausted'' or ``outside the top-$k$ relevance window.''
\end{enumerate}

This cleanly separates two classes of memory with different service-level
requirements: ``always honor'' instructions (a security constraint must never be
dropped because the budget is tight) versus ``recall if relevant'' facts
(competing for the remaining budget by score). The assembled system prompt wraps
the selected memories in an explicit \texttt{ACTIVE LONG-TERM MEMORY} block with
instructions to honor critical items, prefer the most recent decision on
conflict, and never use superseded/expired content, reinforcing at the prompt
level the policy already enforced at the data level.

% ===========================================================================
\section{Non-Destructive Forgetting}
\label{sec:forgetting}

A forgetting sweep runs over the store and applies three rules, each emitting an
auditable timeline event:

\begin{itemize}[leftmargin=1.4em,itemsep=1pt]
  \item \textbf{Expire.} A memory with a past \texttt{expires\_at} transitions
  to \texttt{expired} (e.g. ``the staging server is down today only'').
  \item \textbf{Archive.} An \texttt{active}, non-critical memory unused for more
  than 30 days with importance below $0.35$ transitions to \texttt{archived}.
  \item \textbf{Supersede.} When extraction detects a contradicting decision, the
  old record is linked via \texttt{superseded\_by} and excluded from retrieval.
\end{itemize}

No rule hard-deletes data; deletion happens only on explicit user request.
Expired/archived/superseded records remain queryable on the timeline and in the
Memory Graph but are excluded from context, so the agent never reasons from
invalid memory while the user retains a complete, reversible history. This is the
operational realization of our thesis that forgetting should be \emph{governed
and explainable} rather than lossy.

% ===========================================================================
\section{Transparency: The Memory Trace}
\label{sec:trace}

Every chat response carries a \texttt{MemoryTrace} object recording the complete
decision surface: the list of \emph{included} memories with their full component
breakdown and inclusion reason, the list of \emph{skipped} memories with their
reason, the token budget and tokens used, the number of candidates considered,
and retrieval latency. The trace is rendered in the UI but is also a plain API
object, making the system debuggable and auditable end-to-end. To our knowledge
this degree of built-in inclusion-level explanation is uncommon among
memory-augmented assistants and is, we argue, the practical embodiment of the
interpretability the scoring function was designed to provide.

% ===========================================================================
\section{Evaluation}
\label{sec:eval}

\subsection{Protocol}
We implement a benchmark harness that runs 24 diagnostic scenarios spanning six
capability categories (Table~\ref{tab:scenarios}). Each scenario seeds one or
more \emph{setup messages} (which \osname{} converts to memories), then poses a
\emph{test question}. The harness evaluates two systems on the identical
question:

\begin{itemize}[leftmargin=1.4em,itemsep=1pt]
  \item \textbf{Memory agent.} The question is answered with the \osname{}-built
  context (retrieve $\rightarrow$ score $\rightarrow$ budget assembly).
  \item \textbf{No-memory baseline.} The same model answers with a bare system
  prompt and no long-term memory, isolating the contribution of the memory
  layer.
\end{itemize}

We grade two complementary quantities. \emph{Retrieval-level} metrics
(\textsc{recall@5}, outdated-memory leaks, token savings, latency) are
properties of the \osname{} layer and are independent of the answer model.
\emph{Answer-level} accuracy grades the text the LLM actually produces, and we
report it for each of several interchangeable backbones (\S\ref{sec:crossmodel})
to test whether the benefit of the memory layer generalizes across models.

Correctness is decided by keyword criteria attached to each scenario:
\emph{expected} keywords that must be recalled and \emph{forbidden} keywords that
must not leak. To avoid penalizing legitimate references to a superseded choice
(e.g. ``use Next.js \emph{instead of} React''), a forbidden phrase counts as a
\emph{leak} only when it appears in an injected memory that does not also carry an
expected phrase, i.e. a genuinely outdated memory rather than a superseding
decision. From these we compute task accuracy, \textsc{recall@5}, supersession
success, outdated-memory leaks, an accuracy delta over baseline, and token
savings versus a full-history baseline. The headline numbers are from a live
\textsc{Qwen} backend (\texttt{qwen3.7-max}); the harness also runs against a
deterministic offline backend that reflects the memory \emph{mechanism} in
isolation and is exactly reproducible without credentials, so reported behavior
does not depend on a specific model endpoint.

\begin{table}[t]
\centering
\small
\caption{The six capability categories probed by the 24-scenario benchmark
(four scenarios per category). Together they probe the capabilities a
governance-oriented memory layer must exhibit.}
\label{tab:scenarios}
\begin{tabular}{@{}llp{0.46\linewidth}@{}}
\toprule
\textbf{Scenario} & \textbf{Capability} & \textbf{Probe} \\
\midrule
Preference recall & Intrinsic recall & Recall a stated stack preference. \\
Cross-session recall & Persistence & Recall a project decision made earlier. \\
Contradiction / supersession & Lifecycle & Honor the \emph{new} decision, suppress the old. \\
Expired deadline & Expiry & Do \emph{not} report a temporary fact that has expired. \\
Critical memory recall & Guaranteed inclusion & Always surface a critical constraint. \\
No-memory vs.\ memory & End-to-end & Compose multiple preferences into one answer. \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Results}
Table~\ref{tab:results} reports aggregate metrics from the 24-scenario harness
with a live \textsc{Qwen} backend (\texttt{qwen3.7-max}, accessed through the
Alibaba Cloud DashScope OpenAI-compatible endpoint; \texttt{text-embedding-v3}
embeddings). The memory agent answers $0.75$ of test questions correctly versus
$0.50$ for the no-memory baseline (a $+0.25$ delta), retrieves the target fact
into context at \textsc{recall@5} $=0.77$, and reduces injected context by
$\approx 98\%$ relative to replaying the full history, all within the
$2{,}500$-token budget and at a mean memory-retrieval latency under $5$~ms. The
absolute scores are deliberately conservative: correctness is strict keyword
matching on the generated answer, and the released system strengthens
supersession beyond the configuration measured here (Table~\ref{tab:results} is
a lower bound on lifecycle accuracy).

\begin{table}[t]
\centering
\small
\caption{Benchmark results with a live \textsc{Qwen} backend
(\texttt{qwen3.7-max}; 24 scenarios). The memory layer is the only difference
between the two systems.}
\label{tab:results}
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Metric} & \textbf{Memory agent} & \textbf{No-memory baseline} \\
\midrule
Task accuracy            & \textbf{0.75} & 0.50 \\
Accuracy delta           & \multicolumn{2}{c}{$+0.25$} \\
\textsc{recall@5}        & 0.77 & n/a \\
Outdated-memory avoidance& 0.83 & n/a \\
Token savings vs.\ full history & $\approx 98\%$ & n/a \\
Avg.\ retrieval latency  & $\approx 4$~ms & n/a \\
Context budget           & 2{,}500 tok & (full history) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cross-model generalization}
\label{sec:crossmodel}
Because \osname{} is decoupled from the answer model, we re-run the benchmark
holding the entire memory layer fixed and swapping only the LLM that generates
the final answer. Table~\ref{tab:crossmodel} reports the memory agent and the
no-memory baseline for two frontier backbones. The memory layer improves
accuracy for \emph{both} models, by $+0.25$ for \texttt{qwen3.7-max} and $+0.12$
for \texttt{gpt-4o}. The smaller delta for the stronger general model is
expected, since a more capable model recovers more answers from priors alone;
the consistent positive gain is evidence that the contribution we measure is the
memory layer rather than any single model.

\begin{table}[t]
\centering
\small
\caption{Cross-model generalization. The \osname{} layer (extraction, scoring,
retrieval, budgeting) is identical across rows; only the answer-generating LLM
changes. \textsc{Qwen} is the production model; GPT-4o is included to test
generalization.}
\label{tab:crossmodel}
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Answer backbone} & \textbf{Memory agent} & \textbf{No-memory} & \textbf{Delta} \\
\midrule
Qwen (\texttt{qwen3.7-max}) & \textbf{0.75} & 0.50 & $+0.25$ \\
OpenAI (\texttt{gpt-4o})    & \textbf{0.79} & 0.67 & $+0.12$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Analysis}
The accuracy gap is concentrated in exactly the scenarios the baseline cannot
solve in principle: cross-session recall and supersession require state that a
stateless model does not have, so the baseline's successes come only from
generic priors (e.g. naming a popular framework) rather than from the user's
actual choices. The remaining agent errors are concentrated in supersession,
where a live model paraphrases the user's switch (e.g. ``move from Flask to
FastAPI'') in ways that a purely lexical contradiction rule can miss; the
released system mitigates this with replacement-cue-aware detection and by
honoring the extractor's structured supersede actions, and full elimination of
such leaks is a function of extraction quality rather than of the lifecycle
mechanism itself, which excludes any record once it is marked superseded or
expired. The large token reduction (about $98\%$ on these scenarios) shows that
selective, scored injection under a budget retains the relevant facts while
discarding the bulk of raw history, addressing both cost and the
``lost-in-the-middle'' degradation of long contexts~\citep{liu2024lostmiddle}.

\subsection{Ablation: which mechanism matters?}
\label{sec:ablation}
To isolate the contribution of each governance mechanism we disable one at a
time and measure retrieval/assembly quality at the memory-layer stage, without
answer generation, which makes the study deterministic and exactly reproducible
on the offline backend (Table~\ref{tab:ablation}). The decisive result is that
\emph{lifecycle exclusion} is what suppresses outdated memory: removing it
(treating superseded and expired records as eligible and zeroing their
penalties) raises the outdated-leak rate ninefold, from $0.04$ to $0.38$, with no
recall benefit. By contrast, \textsc{recall@5} and critical inclusion are
insensitive to the exact scoring weights (both unchanged under a similarity-only
ranking and under uniform weights), because on these stores the hybrid
similarity already surfaces the target and the score's critical bonus already
ranks critical memories highly. The quality of the layer therefore comes from
its \emph{governance machinery}, lifecycle exclusion and the explicit penalties,
rather than from delicate weight tuning, which is consistent with our central
thesis; the hard critical-inclusion guarantee (\S\ref{sec:budget}) is a
worst-case safety net on top of the bonus that binds only under extreme budget
pressure.

\begin{table}[t]
\centering
\small
\caption{Governance ablation (24 scenarios, deterministic offline backend).
Disabling lifecycle exclusion alone increases outdated-memory leaks ninefold;
recall and critical inclusion are insensitive to the scoring weights.}
\label{tab:ablation}
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Variant} & \textbf{\textsc{recall@5}} & \textbf{Leak rate} & \textbf{Critical incl.} \\
\midrule
Full (proposed)              & 0.95 & \textbf{0.04} & 1.00 \\
$-$ lifecycle exclusion      & 1.00 & 0.38          & 1.00 \\
similarity-only ranking      & 0.95 & 0.04          & 1.00 \\
uniform weights              & 0.95 & 0.04          & 1.00 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{A standard-benchmark first cut: LoCoMo}
\label{sec:locomo}
To situate the system on a shared, external benchmark we run a first cut on
LoCoMo~\citep{maharana2024locomo}, the long-conversation memory dataset used by
recent production systems. We ingest three conversations \emph{verbatim} (one
memory per dialogue turn, $1{,}451$ turns total), answer all $150$
non-adversarial questions with the standard retrieve\,$\to$\,score\,$\to$\,budget
assembly (top-$k=8$ within the $2{,}500$-token budget) using
\texttt{qwen3.7-max}, and grade with strict token-level F1 and exact match by
category (Table~\ref{tab:locomo}). Overall F1 is $0.252$ and EM $0.093$;
\emph{evidence recall@8}, the model-independent fraction of questions whose
annotated evidence turn was actually injected into the context, is $0.358$
(against a $k/|store|\!\approx\!0.02$ random baseline).

\begin{table}[t]
\centering
\small
\caption{LoCoMo first cut (3 conversations, $1{,}451$ verbatim turns, $150$
questions; \texttt{qwen3.7-max}, top-$k=8$, strict token-F1). The lower block
conditions F1 on whether the evidence turn was retrieved into context.}
\label{tab:locomo}
\begin{tabular}{@{}lcccc@{}}
\toprule
\textbf{Category} & \textbf{n} & \textbf{F1} & \textbf{EM} & \textbf{Ev.\ recall@8} \\
\midrule
Temporal    & 67  & 0.308 & 0.164 & 0.403 \\
Open-domain & 13  & 0.331 & 0.077 & 0.364 \\
Multi-hop   & 57  & 0.189 & 0.018 & 0.333 \\
Single-hop  & 13  & 0.158 & 0.077 & 0.231 \\
\textbf{Overall} & \textbf{150} & \textbf{0.252} & \textbf{0.093} & \textbf{0.358} \\
\midrule
\multicolumn{5}{@{}l}{\emph{Conditioned on evidence retrieval:}} \\
Evidence retrieved & 53  & \textbf{0.572} & 0.245 & n/a \\
Evidence missed    & 95  & 0.068 & 0.011 & n/a \\
\bottomrule
\end{tabular}
\end{table}

The most informative view is the conditional one. When the memory layer places
the annotated evidence turn into the context, answer F1 is $0.572$; when it does
not, F1 collapses to $0.068$, an \textbf{$8.4\times$} gap. This cleanly
separates the two error sources: the answer model is competent \emph{given} the
right memory, and the binding constraint at this configuration is retrieval
\emph{depth}, $k=8$ memories drawn from $\sim\!480$ candidate turns under a
tight token budget. Depth is a tunable dial (raise $k$ or the budget, or use a
stronger embedding model), not a property of the governance mechanism, and the
harness makes the trade-off directly measurable. We emphasize that these are
deliberately conservative numbers, strict token-F1 rather than LLM-judged
grading, short-phrase prompting, and verbatim per-turn ingestion with no
summarization advantage, and are therefore \emph{not} directly comparable to the
LLM-judged LoCoMo figures reported for Mem0~\citep{chhikara2025mem0} and
Zep~\citep{rasmussen2025zep}. The apparatus for a fair head-to-head, the same
answer model and grader for every system, is released with the paper; the
runner is checkpointed so that full-dataset runs are interruption-safe.

% ===========================================================================
\section{Limitations and Future Work}
\label{sec:limitations}

\paragraph{Benchmark scope.} The 24-scenario harness is diagnostic rather than
large-scale; its keyword-based correctness, while precise and reproducible, does
not capture answer fluency or subtle reasoning. The cross-backbone study covers
two frontier models (Qwen and GPT-4o); broadening it to more models and a
larger, multi-session benchmark with human or model-graded answers. Our LoCoMo
first cut (\S\ref{sec:locomo}) is a step in this direction but is deliberately
conservative: strict token-F1 rather than LLM-judged grading, verbatim per-turn
ingestion, and a shallow $k=8$ retrieval budget. The $8.4\times$
evidence-conditioned gap shows retrieval depth, not answering, is the current
bottleneck, so the clearest next steps are a top-$k$/budget sweep, an
LLM-judged grading mode, and a full-dataset, identically-graded comparison
against external systems (Mem0~\citep{chhikara2025mem0},
Zep~\citep{rasmussen2025zep}, MemGPT~\citep{packer2023memgpt},
A-Mem~\citep{xu2025amem}). The released adapter and its checkpointed runner make
these runs reproducible and interruption-safe.

\paragraph{Hand-set weights.} The scoring weights in Equation~\eqref{eq:score}
are interpretable priors, not learned. Learning them, e.g. by optimizing
downstream task accuracy or from preference feedback, while preserving the
linear, inspectable form is an attractive direction, as is per-user adaptation.

\paragraph{Extraction reliability.} Memory quality is bounded by extraction
quality; erroneous or over-eager extraction propagates downstream. Confidence
calibration, extraction-level evaluation, and user-in-the-loop confirmation for
low-confidence or sensitive memories would strengthen robustness.

\paragraph{Token estimation and budgeting.} The $\approx\!4$-characters/token
heuristic is adequate for budgeting but approximate; exact tokenizer-aware
accounting and adaptive budgets per model would tighten the guarantees.

% ===========================================================================
\section{Conclusion}
\label{sec:conclusion}

We presented \sysname{}, a persistent-memory agent built on \osname{}, a memory
layer that treats long-term memory as a \emph{governance} problem rather than a
capacity problem. By making memory typed, scoring it with a fully interpretable
linear model, assembling context under a budget that guarantees critical
inclusion, forgetting through an auditable state machine, and exposing a complete
per-answer trace, the system remembers what matters, refuses to act on what has
become invalid, and can explain every decision. On a reproducible 24-scenario
benchmark it improves task accuracy over a no-memory baseline for two frontier
backbones ($+0.25$ for \texttt{qwen3.7-max}, $+0.12$ for \texttt{gpt-4o}) while
cutting injected context by roughly $98\%$, evidence that the benefit comes from
the memory layer rather than any single model. An ablation shows the layer's
trustworthiness rests on lifecycle governance (removing exclusion raises
outdated-memory leaks ninefold), and an evidence-conditioned LoCoMo analysis
($8.4\times$ F1 gap) localizes the remaining headroom to retrieval depth, a
tunable parameter, rather than to the mechanism. We release the system, the
benchmarks, and the trace format to support reproducible research on
trustworthy agent memory.

% ===========================================================================
\section*{Reproducibility and Availability}
The complete system (backend, frontend, the benchmark harnesses, and the test
suite) runs offline with no external credentials via a deterministic
embedding/LLM fallback. All mechanism-level results, the governance ablation
(Table~\ref{tab:ablation}) and the retrieval diagnostics, are exactly
reproducible on commodity hardware with no API key; the live-model results
(Tables~\ref{tab:results}, \ref{tab:crossmodel}, and \ref{tab:locomo}) are
reproducible given API access to the corresponding models, and the LoCoMo
runner is checkpointed so long runs are interruption-safe. The scoring weights,
scenario definitions, and trace schema are specified in
\S\ref{sec:scoring}--\S\ref{sec:trace} and in the released code.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
